---
title: "DataCompPractice"
author: "Jerad Acosta"
date: "March 6, 2015"
output: pdf_document
---

### Load Data
```{r}
data <- read.table("/Users/irJERAD/1DataContest/ExpulsionSuspensionData/2013-14SuspExpData.txt", sep="\t", header = T, fill = T)
# Use fill = T because starting at line 5016 missing values have an asterisk "*"
```

**Variable Names**
> names(d)
 [1] "AggegateLevel"         "Cds"                   "Name"                  "DisciplineType"       
 [5] "Ethnicity"             "Weapons"               "Drugs"                 "ViolenceWithInjury"   
 [9] "ViolenceWithoutInjury" "OtherNonDefiance"      "OtherDefiance"         "Total"                
[13] "Year"                  "DateCreated"           "DateUpdated"  

### Dimensions (pre Tidy)
Data set has 15 variables and 72310 Observations
```{r}
dim(data)
# [1] 72310    15
sum(as.integer(data$Drugs))
# [1] 604509
# OR sum(as.integer(data$[[7]]))
```
#### All Sums
[[6]] Weapons = 324726   
[[7]] Drugs = 604509   
[[8]] ViolenceWithoutInjury = 611473   
[[9]] ViolenceWithInjury = 2124247   
[[10]] OtherNonDefiance = 598305   
[[11]] OtherDefiance = 2107312   
[[12]] Total = 4950594   

**There is Some Overlap**   
bc 324726 + 604509 + 611473 + 2124247 + 598305 + 2107312 = 6370572   
but total is 4950594

**Need to create Tidy data set**

### Creating a Tidy data subset for analysis
#### By School
```{r}
library(dplyr) # to subset observations and Variables
schoolTotals <- filter(data, AggegateLevel == 'T') # Data.frame of school totals
# 27 observations >> so we must have 27 unique schools
# To Check:
```

```{r, eval=FALSE}
schoolTotals == distinct(schoolTotals)
 [1] TRUE
```

```{r}
# coerse classes to appropriate for visualization and grammar of graphics
for (i in 6:12) {
        # Convert Weapons, Drugs, ViolenceWithInjury, ViolenceWithoutInjury
        # OtherNonDefiance OtherDefiance and Total as integers for graphics
        schoolTotals[[i]] <- as.integer(schoolTotals[[i]])
}
# cast Ethnicity as factor for aesthetic differentiation
schoolTotals$Ethnicity <- as.factor(schoolTotals$Ethnicity)

# create data directory
if(!file.exists("RData")) {
    dir.create("RData")
}
# write new data into data file
write.table(schoolTotals, file = "./RData/schoolTotalSuspExp")
```

### Ideas
* Explore unfair punishment (based on ethnicity)
        + DisciplineType across different Ethnicity
        + keep Offense Category constant (Drugs, Weapons, etc)
        + Use Residency in San Diego School by Race/Ethnicity
http://s3.sandiegodata.org.s3.amazonaws.com/share/data_contest_2015/education/sandi.net/Residency%20by%20Ethnicity%201415%20Flat.xlsx
```{r, qplot}
library(ggplot2)
qplot(DisciplineType, Drugs, data = schoolTotals, color = as.factor(Ethnicity))
# Ethnicity is an integer, need to cast as factor
```

#### rCharts D3 JavaScript Library
```{r, rCharts1, results='asis', comment=NA, tidy=FALSE}
library(devtools)
library(rCharts)
# Create Stacked and grouped interactive Bar Charts
n1 <- nPlot(Weapons ~ DisciplineType, group = "Ethnicity", data = schoolTotals, type = "multiBarChart", cdn = TRUE)
n1$show('inline', include_asset = TRUE)
```

```{r, rCharts2, results='asis', comment=NA, tidy=FALSE}
# Lined Dot
x1 <- xPlot(ViolenceWithoutInjury ~ DisciplineType, group = "Ethnicity", data = schoolTotals, type = "line-dotted", cdn = TRUE)
x1$show('inline', include_assets = TRUE)
```
Notice the drop in "In-School Suspensions" for specific Ethnic Groups stand out vividly when disciplining for violent actions that did not result in any injury.   
One could suspect such disciplinary concequences are very subjective and left to the disciplinarians discretion, which in this case appears to have a strong correlation to student ethnicity.
```{r, rCharts3, results='asis', comment=NA, tidy=FALSE}
# Total Disciplinary Action (Each a different color) Vs Ethnicity
n1 <- nPlot(Total ~ Ethnicity, group = "DisciplineType", data = schoolTotals, type = "multiBarChart", cdn = TRUE)
n1$show('inline', include_assets = TRUE)
```

In the last plot we are looking at the total of each disciplinary action against each Ethnic group.   
**Dark Blue** Represents Expulsion   
**Light Blue** Represents In-School Suspension   
**Orange** Represents Out-of-School Suspension   
**Number to Ethnicity** transcription   

0=Not Reported   
1=American Indian or Alaskan Native, not Hispanic   
2=Asian, not Hispanic   
3=Pacific Islander, not Hispanic 4=Filipino, not Hispanic 5=Hispanic or Latino   
6=African American, not Hispanic 7=White, not Hispanic   
9=Two or more races, not Hispanic   

Ideally we would like to see students spending as much time in school as possible.
Some of the benefits to keeping students in school:
* Education
        + Personal Enrichment and development
        + Social Skills
        + Academic Skills
* Economic
        + Allowing Parents to work
        + Saving money on child care for parents
        + Investing in Students economic future via education
        
Thus, due to the importance of keeping students in school we would want to see Expulsion as a last resort for discipline.
Using similar logic, we can conclude that In-School Suspensions should be considered a better alternative to Out-Of-School Suspensions whenever possible.

Unfortunately the data does not follow this pattern of thought.   

Additionally we can see a great disproportion between disciplinary actions given to students of different ethnicities.
Minorities such as American Indian's are given over 5 times more out of school suspensions than in-school suspensions in total.
The minority groups of Pacific Islander and Filipino as well as the Hispanic or Latino groups each have more than one-third of their disciplinary actions resulting in complete expulsion from school.   
**TODO** 
* Show these groups are also financially disadvantaged and draw a link to the wealth of a group of students and the disciplinary actions they receive.
* Check census data to confirm the number of students receiving these disciplinary actions is much greater than the proportion of them enrolled in the schools.
        + could also check STAR test data for student ethnicity enrollment rates.